---
title: "FunCN tutorial"
output:
  html_document:
    df_print: paged
---
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,    
  message = FALSE,   
  warning = FALSE)
```

```{r, library}
library(ggplot2)
library(dplyr)
library(tibble)
library(cowplot)
```

```{r, load data}
radial_pts <- readRDS('../data/radial_pts.rds')
equidist_pts <- readRDS('../data/equidist_pts.rds')
```

# Case 1: Equidistance example
```{r, equi-distance dataset}
p1 <- ggplot() +
  geom_segment(data = equidist_pts %>% dplyr::filter(label != "A"),
               aes(x = cx, y = cy, xend = x, yend = y, color = label),
               linewidth = 0.7, alpha = 0.7) +
  geom_point(data = equidist_pts, aes(x, y, color = label, shape = label), size = 3.5) +
  geom_point(
    data = radial_pts %>% dplyr::filter(label == "A"),
    aes(x, y, color = label, shape = label),
    size = 3.5, show.legend = FALSE
  ) +
  coord_fixed(xlim = c(0, 50), ylim = c(0, 50)) +
  labs(title = "(i)Equal distance", x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.title = element_blank(), 
        plot.title = element_text(hjust = 0.5))
```

# Case 2: Radial-distance example 
```{r, radial}
# Parameters
cx <- 25; cy <- 25
r_B <- 9
r_C <- 17
r_D <- 25

# circles for B, C, D 
circle_df <- function(cx, cy, r, n = 360) {
  th <- seq(0, 2*pi, length.out = n)
  tibble(x = cx + r*cos(th), y = cy + r*sin(th))
}

circles <- bind_rows(
  circle_df(cx, cy, r_B) %>% mutate(label = "B"),
  circle_df(cx, cy, r_C) %>% mutate(label = "C"),
  circle_df(cx, cy, r_D) %>% mutate(label = "D")
)

color_map <- c(B = "#1b9e77", C = "#d95f02", D = "#7570b3")

# plot
p2 <- ggplot() +
 geom_path(data = circles, aes(x, y, color = label), linewidth = 0.8, linetype = 2) +
  geom_segment(data = radial_pts %>% dplyr::filter(label != "A"),
               aes(x = cx, y = cy, xend = x, yend = y, color = label),
               linewidth = 0.7, alpha = 0.7) +
  geom_point(data = radial_pts, aes(x, y, color = label, shape = label), size = 3.5) +
  # draw A on top
  geom_point(
    data = radial_pts %>% dplyr::filter(label == "A"),
    aes(x, y, color = label, shape = label),
    size = 3.5, show.legend = FALSE
  ) +
  coord_fixed(xlim = c(0, 50), ylim = c(0, 50)) +
  labs(title = "(ii) Radial distance", x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.title = element_blank(), 
        plot.title = element_text(hjust = 0.5))
```

```{r, plot}
plot_grid(p1, p2, nrow = 1)
```

# Apply FunCN function to two datasets 
**Description**  
Computes smoothed, leave-one-out kernel weights for each point, grouped by cell-type marks.

**Requirements**
Input `expr` must contain:
  
  - `x_col`: x-coordinates  
  - `y_col`: y-coordinates  
  - `celltype_col`: cell type (mark)

**Multi-sample data**  
If there is multi-sample data in one data frame:
  
  - Set `multisample = TRUE`  
  - Add `sample_col = "sample_col_name"`

**Smoothing kernels**  

  - A character string specifying the kernel  
  - spatstat smooth function Options: `"gaussian"`, `"epanechnikov"`, `"quartic"`, `"disc"`
  
```{r, apply FunCN}
source('./functions/do_CI_quantification.R') 

# check data format 
print(radial_pts)

radial_pts <- as.data.frame(radial_pts) # transform to data.frame class 
equidist_pts <- as.data.frame(equidist_pts) 

```

# Usage
```{r, usage}
equi_spatwt <- do_CI_quantification(expr = equidist_pts,
                                    x_col = "x",
                                    y_col = "y",
                                    celltype_col = "label",
                                    multisample = FALSE, 
                                    sample_col = NA, 
                                    kernels = "gaussian",
                                    sigma = 10)

radi_spatwt <- do_CI_quantification(expr = radial_pts,
                                    x_col = "x",
                                    y_col = "y",
                                    celltype_col = "label",
                                    multisample = FALSE, 
                                    sample_col = NA, 
                                    kernels = "gaussian",
                                    sigma = 10)
print(equi_spatwt)
print(radi_spatwt)
```

