---
title: "FunCN (Functional Cellular Neighborhood) tutorial"
output:
  html_document:
    df_print: paged
---
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,    
  message = FALSE,   
  warning = FALSE)
```

# Introduction
**Why we developed FunCN?**  
We developed **FunCN** to overcome the limitations of conventional neighborhood analysis methods such as KNN approach. The traditional methods rely on fixed boundaries, assigning equal weights to all neighboring cells within a cutoff distance, which can obscure the nuanced influence of spatial proximity. In reality, cellular interactions are distance-dependent, where closer cells exert stronger biological effects than those farther away. To address this, **FunCN** quantifies cellular influence through a continuous, distance-weighted representation that captures the gradual decay of interaction strength across spatial scales. 

**RESOURCES**
Cho, Y., Lee, J.W., Shin, S.M., Hernandez, A.G., Yuan, X., Schneider, J., Hooper, J.E., Wood, L.D., Jaffee, E.M., Deshpande, A. and Ho, W.J., 2025. Modeling cellular influence delineates functionally relevant cellular neighborhoods in primary and metastatic pancreatic ductal adenocarcinoma. bioRxiv, pp.2025-06. doi: https://doi.org/10.1101/2025.06.12.659314

# Apply FunCN function
**Description**  
Computes smoothed, leave-one-out kernel weights for each point, grouped by cell-type marks.

**Requirements**
Input `expr` must contain:
  
  - `x_col`: x-coordinates  
  - `y_col`: y-coordinates  
  - `celltype_col`: cell type (mark)

**Multi-sample data**  
If there is multi-sample data in one data frame:
  
  - Set `multisample = TRUE`  
  - Add `sample_col = "sample_col_name"`

**Smoothing kernels**  

  - A character string specifying the kernel  
  - spatstat smooth function Options: `"gaussian"`, `"epanechnikov"`, `"quartic"`, `"disc"`

# Load data
```{r, library}
library(ggplot2)
library(dplyr)
library(tibble)
library(cowplot)
library(FNN)
library(tidyr)
```

```{r, load data}
radial_pts <- readRDS('../data/radial_pts.rds')
equidist_pts <- readRDS('../data/equidist_pts.rds')
```

```{r, apply FunCN}
source('./functions/compute_FunCN.R') 

# check data format 
print(radial_pts)

radial_pts <- as.data.frame(radial_pts) # transform to data.frame class 
equidist_pts <- as.data.frame(equidist_pts) 
```

# Usage
```{r, usage}
equi_spatwt <- compute_FunCN(expr = equidist_pts,
                             x_col = "x",
                             y_col = "y",
                             celltype_col = "label",
                             multisample = FALSE, 
                             sample_col = NA, 
                             kernels = "gaussian",
                             sigma = 10)

radi_spatwt <- compute_FunCN(expr = radial_pts,
                             x_col = "x",
                             y_col = "y",
                             celltype_col = "label",
                             multisample = FALSE, 
                             sample_col = NA, 
                             kernels = "gaussian",
                             sigma = 10)

print(equi_spatwt)
print(radi_spatwt)
```

# Compare the FunCN score with KNN
```{r, bar plot for FunCN score}
get_neighbor_props <- function(df, k = 10, label_col = "label", xy_cols = c("x","y"), all_labels = NULL) {
  # run kNN
  knn_idx <- get.knn(df[, xy_cols], k = k)$nn.index
  # pull labels as vector
  labels_vec <- df[[label_col]]
  # map neighbor indices -> labels
  knn_labels <- matrix(labels_vec[knn_idx],
                       nrow = nrow(knn_idx),
                       ncol = ncol(knn_idx))
  
  # count label frequency per row
  freq_mat <- t(apply(knn_labels, 1, function(neigh_labels) {
    tab <- table(factor(neigh_labels, levels = all_labels))
    as.numeric(tab)
  }))
  colnames(freq_mat) <- all_labels
  
  # convert to proportions
  prop_mat <- as.data.frame(freq_mat/k)
  # set rownames: "<rownames>:<label>"
  rownames(prop_mat) <- paste0(rownames(df), ":", df[[label_col]])
  return(prop_mat)
}

# define labels
all_labels <- unique(c(equidist_pts$label, radial_pts$label))

# run for each dataset
prop_df_equi <- get_neighbor_props(equidist_pts, k = 10, all_labels = all_labels)
prop_df_radi <- get_neighbor_props(radial_pts,k = 10, all_labels = all_labels)

print(prop_df_equi)
print(prop_df_radi)
```

# Compare spatial organization with cellular influence score

### Case 1: Equidistance example
```{r, equi-distance dataset}
p1 <- ggplot() +
  geom_segment(data = equidist_pts %>% dplyr::filter(label != "A"),
               aes(x = cx, y = cy, xend = x, yend = y, color = label),
               linewidth = 0.7, alpha = 0.7) +
  geom_point(data = equidist_pts, aes(x, y, color = label, shape = label), size = 3.5) +
  geom_point(
    data = radial_pts %>% dplyr::filter(label == "A"),
    aes(x, y, color = label, shape = label),
    size = 5, show.legend = FALSE
  ) +
  coord_fixed(xlim = c(0, 50), ylim = c(0, 50)) +
  labs(title = "(i) Equal distance", x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.title = element_blank(), 
        plot.title = element_text(hjust = 0.5))+
  scale_shape_manual(
    values = c("A" = 18,"B" = 17, "C" = 15, "D" = 16))  
```

### Case 2: Radial-distance example 
```{r, radial}
# Parameters
cx <- 25; cy <- 25
r_B <- 9
r_C <- 17
r_D <- 25

# circles for B, C, D 
circle_df <- function(cx, cy, r, n = 360) {
  th <- seq(0, 2*pi, length.out = n)
  tibble(x = cx + r*cos(th), y = cy + r*sin(th))
}

circles <- bind_rows(
  circle_df(cx, cy, r_B) %>% mutate(label = "B"),
  circle_df(cx, cy, r_C) %>% mutate(label = "C"),
  circle_df(cx, cy, r_D) %>% mutate(label = "D")
)

color_map <- c(B = "#1b9e77", C = "#d95f02", D = "#7570b3")

# plot
p2 <- ggplot() +
 geom_path(data = circles, aes(x, y, color = label), linewidth = 0.8, linetype = 2) +
  geom_segment(data = radial_pts %>% dplyr::filter(label != "A"),
               aes(x = cx, y = cy, xend = x, yend = y, color = label),
               linewidth = 0.7, alpha = 0.7) +
  geom_point(data = radial_pts, aes(x, y, color = label, shape = label), size = 3.5) +
  # draw A on top
  geom_point(
    data = radial_pts %>% dplyr::filter(label == "A"),
    aes(x, y, color = label, shape = label),
    size = 5, show.legend = FALSE
  ) +
  coord_fixed(xlim = c(0, 50), ylim = c(0, 50)) +
  labs(title = "(ii) Radial distance", x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.title = element_blank(), 
        plot.title = element_text(hjust = 0.5))+
  scale_shape_manual(
    values = c("A" = 18,"B" = 17, "C" = 15, "D" = 16))   
```
### Compile data for cellular influence score
```{r, plot}
# extract scores for A (reference)
to_long <- function(df, method_name, sample_name) {
  df %>%
    as.data.frame() %>%
    pivot_longer(
      cols = c("A","B","C","D"),
      names_to = "label",
      values_to = "value"
    ) %>%
    mutate(
      method = method_name,
      sample = sample_name
    )
}

long_equi_fun <- to_long(equi_spatwt[1, ], method_name = "FunCN", sample_name = "equidist")
long_radi_fun <- to_long(radi_spatwt[1, ], method_name = "FunCN", sample_name = "radial")
long_equi_knn <- to_long(prop_df_equi[1, ], method_name = "KNN", sample_name = "equidist")
long_radi_knn <- to_long(prop_df_radi[1, ], method_name = "KNN", sample_name = "radial")

data_for_plot <- as.data.frame(rbind(long_equi_fun, long_radi_fun, long_equi_knn, long_radi_knn))

p3 <- ggplot(data_for_plot[data_for_plot$sample=="equidist", ], 
             aes(x=label, y=value, fill=method))+
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = round(value, 3)),              
    position = position_dodge(width = 0.9),   
    vjust = -0.3,                             
    size = 3
  ) +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  labs(x = NULL, y = "Scores", title = NULL)

p4<- ggplot(data_for_plot[data_for_plot$sample=="radial", ], 
             aes(x=label, y=value, fill=method))+
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = round(value, 3)),              
    position = position_dodge(width = 0.9),   
    vjust = -0.3,                             
    size = 3
  ) +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  labs(x = NULL, y = "Scores", title = NULL)

```

```{r fig.width=10, fig.height=10}
plot_grid(p1, p2, p3, p4, ncol = 2, rel_heights = c(4, 1))
```
